Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-01-29 paper/paper.tex
paper: Add State of the Art

Normally, the characterization process takes into account astrometric features
such as proper motion in right ascension and declination, and parallax.
And photometric features like $E(b-r) / G_{mag}$.
Photometric variables are used for generating the H-R diagram of the OC candidate stars.
This stars should present a sharp profile corresponding to an isochrone curve derived
from a theoretical model.
This model does consider metallicity, mass and brightness of the stars involved.
At the moment, a set of tools have been used for this purpose, as explained later,
but in general, these tools require supervised and parameterized models.
So a previous knowledge of the cluster is necessary or the process must be repeated
iteratively in order to achieve valid results.

Our aim is to build a machine learning model capable of characterizing open
clusters inside a stellar region without previous knowledge of that region.
This model will take advantage of these features to train an Artificial Neural
Network (ANN) that will clusterize all stars within the given region.
One of these groups will be the open cluster that interests us.
With that purpose we have adapted the Unsupervised Deep Embedding for Clustering
Analysis \cite{xie2016unsupervised} to our problem.

This model fulfills one of our main requirements.
It has to be \emph{non-supervised} and \emph{non-parameterized} in order to make easier
the automation process of analyzing a wide range of regions with different topologies.
Another important requirement for our model is that it must be computationally efficient
in order to run in common workstations.
The model has been developed with Python using the Keras framework which takes
advantage of modern GPUs to make its computations.
That increases significantly the computational capacity of our model making possible
to run it on regular workstations with one of those GPUs.

@@

As shown in Section \ref{sec:results}, the results obtained are promising.
Although some improvements must be done in order to consider it as valid tool
that could be part of the standard set of tools for open clusters characterization.

@@

Estudio a fondo el dominio de aplicación, citando numerosas referencias.
Debe aportar un buen resumen del conocimiento que ya existe en el campo de los problemas habituales identificados.
\section{AIMS AND METHOD}

>>>>

Normally, the characterization process takes into account astrometric features
such as proper motion in right ascension and declination, and parallax.
Photometric features like $E(b-r) / G_{mag}$ are considered too.
These last variables are used for generating the H-R diagram of the OC candidate stars.
This stars should present a sharp profile corresponding to an isochrone curve derived
from a theoretical model.
This model does consider metallicity, mass and brightness of the stars involved.
At the moment, a set of tools have been used for this purpose, as explained later,
but in general, these tools require supervised and parameterized models.
So a previous knowledge of the cluster is necessary or the process must be repeated
iteratively in order to achieve valid results.

Our aim is to build a machine learning model capable of characterizing an open
cluster inside a stellar region with knowing nothing about the region.
This model will take advantage of these features to train an Artificial Neural
Network (ANN) that will clusterize all stars within the given region.
One of these groups will be the open cluster that interests us.
With that purpose we have adapted the Unsupervised Deep Embedding for Clustering
Analysis \cite{xie2016unsupervised} to our problem.

This model fulfills one of our main requirements.
It has to be \emph{non-supervised} and \emph{non-parameterized} in order to make easier
the automation process of analyzing a wide range of regions with different typologies.
Another important requirement for our model is that it must be computationally efficient
in order to run in common workstations.
The model has been developed with Python using the Keras framework which takes
advantage of modern GPUs to make its computations.
That increases significantly the computational capacity of our model making possible
to run it on regular workstations with one of those GPUs.

@@

As shown in Section \ref{sec:results}, the results obtained are promising.
This allows us to consider our model as a valid tool that could be part
of the standard set of tools for open clusters characterization.

@@

The initial approach to find open clusters is by searching for overdensities
in the astrometric space of the galactic disk and a later identification of
possible OCs using photometric information \cite{castro2020hunting}.
However, although it looks simple, it is not. Here we face the problem that the
near field around the OC is composed of two kind of populations.
The first kind belongs to the membership stars of the OC (from tens or hundreds
to a few thousand) while the second one is composed by a background of stars
that do not belong to the OC (from tens to hundreds of thousands).
To find out which stars do belong to the OC is the problem we try to solve in
this work as this selection is critical to properly characterize the
fundamental properties of the cluster (total mass, age, chemical composition, etc.).

Since the components of the open cluster are born from the same gas cloud and they
share both dynamic properties and relative distance, as well as age and chemical
composition, the first step is to analyze the proper motion configuration space of
the region of interest. This approach is the one used by many tools from the Virtual
Observatory such as TOPCAT \cite{taylor2005topcat}, Clusterix 2.0 \cite{balaguer2020clusterix},
Aladin \cite{bonnarel2000aladin} or VOSA \cite{bayo2008vosa}.

TOPCAT by itself cannot find members of the open cluster. It requires to parameterize
some of the properties of the cluster, what requires a previous knowledge of the cluster.
Or to make a supervised and manual selection of groups based on overdensities
in the proper motion configuration space. Clusterix 2.0, an interactive web-based tool,
can help to this process. It takes the proper motion diagram without making any prior
assumption about the membership of the candidate star and determines empirically
the frequency functions. Clusterix uses normal Gaussian kernel functions, defined as:

\begin{equation*}
  K(a, b) = \frac{1}{2 \pi h^{2}} \exp{ \left[ - \frac{1}{2}\frac{\left( a - a_{i} \right)^{2} + \left( b - b_{i} \right)^{2}}{ h^{2}} \right]}
\end{equation*}

$\left( a, b \right)$ are referred to the proper motion configuration space,
while a point located in the center of the cell $\left( i, j \right)$
provides the maximum contribution for calculating the local density.
Number $h$ is called the \emph{smoothing} parameter and it is measured
in the same units as proper motion.

Clusterix is defined as an unsupervised and non-parameterized tool, but it
critically depends on the selection of three radii.

\begin{itemize}
  \item \textbf{Inner radius} which defines a region with stars that belong to the OC and field stars.
  \item \textbf{Void region}. This region is excluded from the analysis since it is supposed
                to be a transition region with stars that could belong the OC.
  \item \textbf{External region} or field region.
                This is the outer region with stars that do not belong the open cluster.
\end{itemize}

If the analysis performed by Clusterix is successful, it returns the probability
associated to each star in the dataset to belong to the open cluster.
These results can be imported then into TOPCAT to continue the refining process to
get a valid characterization.

Another recent approach to solve this problem based on machine learning techniques
makes a systematic search for overdensities in the astrometric space of the galactic
disk and a subsequent identification of open clusters using photometric information
\cite{castro2020hunting}.

This last method includes two phases: the first one uses an unsupervised clustering
algorithm, DBSCAN, to search for overdensities $(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$,
and then applies a deep learning ANN, previously trained with magnitude diagrams,
to identify isochrone patterns within the detected overdensities and thus proceed
to confirm them as OC.

It should be noted that for the execution of this method, MareNostrum 42
(Barcelona Supercomputing Center) was used.
So the neural network could handle the image recognition process with
isochrone patterns and not applying theoretical models derived from
values such as metallicity or masses, among other.

This work concludes with the identification of 582 new open clusters
distributed along the galactic disk for a galactic declination bellow 20 degrees.
This result increases the number of known open clusters by 45\%.

As told before, our aim in this work, linked to computational limitations compared
to MareNostrum, is not to do a blind search for clusters, but to obtain a method,
in the machine learning environment, that allows the characterization of open clusters.
We require our method to be unsupervised and non-parameterized making it suitable for
automated processes.

We use the Clusterix+TOPCAT method applied to the same clusters that we have analyzed
with our method to validate our results.

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to include it.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission
(located at L2, 1.5 million kilometers from Earth) and operational since 2014.
The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.

As we will explain later, what we need to accomplish our aim is an algorithm
that manages to make data groups based on the dynamic properties of the stars.

There are several clustering algorithms: KMeans, Mean-Shift Clustering, DBSCAN \cite{ester1996density}
among other. While each one behaves better according to the distribution of the objects
to clusterize, we have chosen KMeans by its simplicity and good results.

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the found cluster contains to many outliers.
For that reason, we have search for a KMeans refinement based on an artificial neural network.

The \emph{Unsupervised Deep Embedding for Clustering Analysis} model
or \emph{DEC} \cite{xie2016unsupervised} takes KMeans as its starting point,
but then, it trains an autoencoder to reduce the feature space
and pass this transformed data through a Clustering Layer which refines the previous selection.

\section{AIMS AND METHOD}
